Factors Influencing Abundance & Virus Isolation in Mosquitoes

### 2024-04-26 R/analysis-code/analysis.qmd
updating manuscript

```{r}
library(here)
library(dplyr)
library(lubridate)
library(readr)
library(stargazer)
library(mgcv)
library(tidyr)
library(caret)
library(randomForest)
library(tidymodels)
library(ranger)
library(dials)
library(vip)
library(ggplot2)
library(boot)
library(yardstick)
library(glmnet) # For LASSO
library(dials)
library(mlr3)
library(data.table)
library(Matrix)
library(gbm)
```

@@

>>>>

```{r}
library(here)
library(dplyr)
library(lubridate)
library(readr)
library(stargazer)
library(mgcv)
library(tidyr)
library(caret)
library(randomForest)
library(tidymodels)
library(ranger)
library(dials)
library(vip)
library(ggplot2)
library(boot)
library(yardstick)
library(glmnet) # For LASSO
library(dials)
library(mlr3)
library(data.table)
library(Matrix)
library(gbm)
library(knitr)
library(gt)
```

@@

# Use gt to create a styled table
gt_table <- gt(rmse_table) %>%
  tab_header(
    title = "RMSE Comparison of Models",
    subtitle = "Sorted by Test RMSE in descending order"
  ) %>%
  cols_label(
    Model = "Model",
    Train_RMSE = "Training RMSE",
    Test_RMSE = "Testing RMSE"
  )

# Define the file path using 'here'
png_path <- here("results", "tables", "rmse_table.png")

# Save the gt table as a PNG
gtsave(gt_table, filename = png_path)
```

### 2024-04-26 products/manuscript/Manuscript.qmd
updating manuscript

## Questions/Hypotheses to be addressed
**Which environmental,ecological, and temporal factors influence the
transmission cycles for EEE in southeastern Massachusetts and which
models best describe the relationship between these factors?** 
From prior
research we know that the mosquito lifecyle is influenced by a number
of factors. Temperature plays a major role in the mosquito life cyle
virus transmission dynamics. As temperature increases, the rate of
larval development also increases and the time between when a mosquito
ingests virus in a bloodmeal and when it can transmit it decreases. The
result is that the mosquito population can grow at faster rate and they
will have more time to transmit the virus. However, MEL larvae develop
in well-insulated crypts that are less sensitive to ambient air
temperature changes, so the effect of temperature changes on MEL
development speed may be less pronounced than other vector species that
develop in open, standing water, like ditches and containers.
Precipitation increases the amount of standing water and increases the
chances of larval survival during the mosquito season and also during
the winter months for those species that overwinter in the larval stage.
For EEE, the main enzootic vector, MEL, overwinters as larvae.

@@

## Exploratory/Descriptive analysis
One factor to consider is if early virus isolations are associated with
cases in humans and other mammals. The table below shows the date of the
first EEE virus isolation in mosquitoes and the number of human of other
animal cases by year.

@@

## Basic statistical analysis
Basic linear models assessed environmental factors' impact on the EEE virus transmission cycle. Initial models investigated the link between virus isolation in mosquitoes and weather conditions (average temperature by week and total precipitation). The outcome was whether at least one positive test occurred for a pool in a given week. The predictor in the first model was the average maximum daily temperature from the previous week. The logistic model revealed a significant relationship. Higher temperatures were positively associated with the likelihood of a positive result (odds ratio ≈ 1.069 for each unit increase in temperature; Estimate = 0.06695, SE = 0.02033, p < 0.001). Incorporating Lagged_Avg_TMAX as a predictor improved the model, reducing residual deviance to 405.79 from the null deviance of 417.24, indicating a better fit. However, residual deviance on 299 degrees of freedom suggests unexplained variability, indicating other influential factors. The model's AIC of 411.59 suggests a balance between complexity and predictive power, supporting its utility. Yet, this relationship warrants closer examination due to likely autocorrelation of temperature with progressing weeks during summer months

@@

## Statistical analysis
Before any models were assessed, the RSME for the null model was calculated (2.40). This was used as the basis for assessing model improvement.

Analysis began with a Generalized Additive Model (GAM) with predictors that were selected based on the exploratory single variable GAM. The variables were chosen if their p=values were <0.05. Eight variables met that condition including: the number of hours below 50°F the week before the mosquito collection week (Hrs_Below_50_lag), the number of hours below 50°F the week of collection (Hrs_Below_50_current), the drought severity index for East Coast FEMA regions (DSCI_fema), the average hourly relative humidity the week of collection (Avg_HourlyRelativeHumidity_current), the average hourly windspeed the week of collection (Avg_HourlyWindSpeed_current), the average number of daylight hours per day (average_daylight_hours), the average number of MEL captured by trap (Average_MEL), and the proportion of MEL captured out of the total number of mosquitoes trapped (MEL_Prop).

@@

```{r}
# Define the file path using here()
file_path_rds <- here("results", "tables", "rmse_table.rds")

# Read the RDS file
rmse_table <- readRDS(file_path_rds)
rmse_table
```
# Discussion
The null model RMSE was 2.40. All models showed improvement over the null model on both the training and testing datasets. The GAM shows significant improvement over the null model for the training set, but only slightly improves over the null for the testing set. This suggests that beyond capturing the mean response, this GAM is not well generalized. The substantial improvement in the training set RSME for GBM and Random Forest coupled with less improvement over the null for the testing set, suggest a tendency for overfitting. These models may be capturing the training data's patterns that do not gerneralize well to unseen data. The the gap between the training and testing RMSE for the LASSO and GAMS models suggest there is less overfitting and are slightly better at generalizing than the others. However, given the non-linear relationships, the LASSO model is likely not appropriate for this analysis.

@@

Likely the biggest limitation 
## Conclusions 
While all models demonstrated improvement over null model, none of them were robust enough to be used for reliable prediction. In the next phases of this project, the data will be expanded to include more years and cover a larger portion of Massachusetts. In additon to increasing the the number of observations, other sources of data will be incorporated. Including virus activity from surrounding regions and areas from which birds migrate to Massachusetts.

>>>>

## Questions/Hypotheses to be addressed

**Which environmental,ecological, and temporal factors influence the
transmission cycles for EEE in southeastern Massachusetts and which
models best describe the relationship between these factors?** From
prior research we know that the mosquito lifecyle is influenced by a
number of factors. Temperature plays a major role in the mosquito life
cyle virus transmission dynamics. As temperature increases, the rate of
larval development also increases and the time between when a mosquito
ingests virus in a bloodmeal and when it can transmit it decreases. The
result is that the mosquito population can grow at faster rate and they
will have more time to transmit the virus. However, MEL larvae develop
in well-insulated crypts that are less sensitive to ambient air
temperature changes, so the effect of temperature changes on MEL
development speed may be less pronounced than other vector species that
develop in open, standing water, like ditches and containers.
Precipitation increases the amount of standing water and increases the
chances of larval survival during the mosquito season and also during
the winter months for those species that overwinter in the larval stage.
For EEE, the main enzootic vector, MEL, overwinters as larvae.

@@

## Exploratory/Descriptive analysis

One factor to consider is if early virus isolations are associated with
cases in humans and other mammals. The table below shows the date of the
first EEE virus isolation in mosquitoes and the number of human of other
animal cases by year.

@@

## Basic statistical analysis

Basic linear models assessed environmental factors' impact on the EEE
virus transmission cycle. Initial models investigated the link between
virus isolation in mosquitoes and weather conditions (average
temperature by week and total precipitation). The outcome was whether at
least one positive test occurred for a pool in a given week. The
predictor in the first model was the average maximum daily temperature
from the previous week. The logistic model revealed a significant
relationship. Higher temperatures were positively associated with the
likelihood of a positive result (odds ratio ≈ 1.069 for each unit
increase in temperature; Estimate = 0.06695, SE = 0.02033, p \< 0.001).
Incorporating Lagged_Avg_TMAX as a predictor improved the model,
reducing residual deviance to 405.79 from the null deviance of 417.24,
indicating a better fit. However, residual deviance on 299 degrees of
freedom suggests unexplained variability, indicating other influential
factors. The model's AIC of 411.59 suggests a balance between complexity
and predictive power, supporting its utility. Yet, this relationship
warrants closer examination due to likely autocorrelation of temperature
with progressing weeks during summer months

@@

## Statistical analysis

Before any models were assessed, the RSME for the null model was
calculated (2.40). This was used as the basis for assessing model
improvement.

Analysis began with a Generalized Additive Model (GAM) with predictors
that were selected based on the exploratory single variable GAM. The
variables were chosen if their p=values were \<0.05. Eight variables met
that condition including: the number of hours below 50°F the week before
the mosquito collection week (Hrs_Below_50_lag), the number of hours
below 50°F the week of collection (Hrs_Below_50_current), the drought
severity index for East Coast FEMA regions (DSCI_fema), the average
hourly relative humidity the week of collection
(Avg_HourlyRelativeHumidity_current), the average hourly windspeed the
week of collection (Avg_HourlyWindSpeed_current), the average number of
daylight hours per day (average_daylight_hours), the average number of
MEL captured by trap (Average_MEL), and the proportion of MEL captured
out of the total number of mosquitoes trapped (MEL_Prop).

@@

![](images/rmse_table-01.png)

# Discussion

The null model RMSE was 2.40. All models showed improvement over the
null model on both the training and testing datasets. The GAM shows
significant improvement over the null model for the training set, but
only slightly improves over the null for the testing set. This suggests
that beyond capturing the mean response, this GAM is not well
generalized. The substantial improvement in the training set RSME for
GBM and Random Forest coupled with less improvement over the null for
the testing set, suggest a tendency for overfitting. These models may be
capturing the training data's patterns that do not gerneralize well to
unseen data. The the gap between the training and testing RMSE for the
LASSO and GAMS models suggest there is less overfitting and are slightly
better at generalizing than the others. However, given the non-linear
relationships, the LASSO model is likely not appropriate for this
analysis.

@@

Likely the biggest limitation \## Conclusions While all models
demonstrated improvement over null model, none of them were robust
enough to be used for reliable prediction. In the next phases of this
project, the data will be expanded to include more years and cover a
larger portion of Massachusetts. In additon to increasing the the number
of observations, other sources of data will be incorporated. Including
virus activity from surrounding regions and areas from which birds
migrate to Massachusetts.
